%%%%%%%% ICML 2026 SUBMISSION %%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}

\usepackage{hyperref}

\newcommand{\theHalgorithm}{\arabic{algorithm}}

\usepackage{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

\usepackage[capitalize,noabbrev]{cleveref}
\crefname{proposition}{Proposition}{Propositions}
\Crefname{proposition}{Proposition}{Propositions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\usepackage[textsize=tiny]{todonotes}

\icmltitlerunning{Selective Adversarial Causal Oversight}

\begin{document}

\twocolumn[
  \icmltitle{Selective Adversarial Causal Oversight:\texorpdfstring{\\}{}%
    An Information-Partitioned Benchmark Prototype and Reference Verifier}

  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Jingbei Niu}{zju}
    \icmlauthor{Zeyuan Tong}{zju}
    \icmlauthor{Yuhang Li}{zju}
  \end{icmlauthorlist}

  \icmlaffiliation{zju}{Zhejiang University, Hangzhou, China}

  \icmlcorrespondingauthor{Yuhang Li}{liyuhangnice666@gmail.com}

  \icmlkeywords{Causal Inference, Large Language Models, Selective Prediction,
    Adversarial Evaluation, Benchmarks, Wise Refusal}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
LLMs are increasingly used to evaluate causal claims, but real causal oversight
is both \emph{selective} and \emph{adversarial}: claimants may hold private
information, exploit non-identifiability, and present persuasive but incomplete
arguments. We formalize this setting as \emph{selective adversarial causal
oversight}, where a verifier with only public evidence must decide whether a
natural-language claim is \textsc{valid}, \textsc{invalid}, or
\textsc{unidentifiable}. We present a benchmark contract and prototype
synthetic core with schema-level public/gold partitioning, twelve graph
families across Pearl's hierarchy, structurally grounded attacks, a persuasion
overlay, and OOD buckets for graph, mechanism, attack-family, context-shift,
and paired-flip shifts. We also propose \emph{countermodel-grounded selective
verification}, a reference pipeline that parses the claim, builds an
assumption ledger, searches for observationally compatible countermodels, and
returns an abstention-aware verdict. On a fixed-code exploratory snapshot
(three seeds, ten samples per family, difficulty $0.55$), this reference
verifier exhibits a more favorable accuracy/unsafe-acceptance/wise-refusal
tradeoff than public-view heuristic baselines, while a deliberately leaking
positive control shows how oracle information can inflate benchmark scores. We
also run a five-model API-backed baseline matrix (Qwen2.5 7B/14B/72B, GPT-5.5
xhigh, and DeepSeek v4 max-thinking) with raw-response logging and fallback
rejection. These results are diagnostic evidence for the public/gold
partitioning contract; larger synthetic runs, real-grounded cases, and human
audit remain required before claiming a mature benchmark release.
\end{abstract}

\section{Introduction}
\label{sec:intro}

LLMs are now used as judges, debate moderators, and policy explainers, but
evaluating their \emph{causal} reasoning remains awkward. Existing benchmarks
largely treat causal questions as graded multiple-choice problems
\cite{jin2023cladder,jin2024corr2cause,frohberg2022crass}: the evaluator
inspects an oracle structural causal model, fixes a yes/no answer, and scores
the LLM's prediction against it. This format is convenient but masks two
features that dominate causal oversight in practice.

\textbf{Information asymmetry.} A claimant -- a clinician, policy advocate,
research summarizer, or deceptive agent -- often has access to private
information that the verifier does not: the data-generating process, hidden
confounders, selection mechanisms, or simply domain familiarity. The verifier
sees only public evidence and the claim text. Ignoring this asymmetry
collapses a hard \emph{oversight} problem into an easy
\emph{closed-world recall} problem.

\textbf{Wise refusal.} A core lesson of causal inference is that, given a fixed
public evidence base, many queries are not identifiable
\cite{pearl2009causality,bareinboim2022pearl}. In such cases the correct
behaviour is not to commit to a verdict, but to abstain with a reason. Modern
LLM judges, however, are trained to commit and exhibit sycophancy and
position bias when uncertain
\cite{sharma2024sycophancy,chen2024judge,ye2024justice}. Existing causal
benchmarks rarely measure ``did the model refuse for the right reason'' as a
first-class quantity.

\paragraph{Our task.} We formalize \emph{selective adversarial causal oversight}
(Section~\ref{sec:task}). An attacker generates a natural-language claim with
full knowledge of the gold structural causal model (SCM); a verifier sees only
the public projection of the scenario plus the claim text; and the evaluator
scores the verifier's \emph{three-way} verdict
$\{\textsc{valid}, \textsc{invalid}, \textsc{unidentifiable}\}$ together with a
structured assumption ledger, refusal reason, and witness.

\paragraph{Our benchmark contract.} We construct a prototype synthetic core of
twelve graph families that explicitly mark family-level identifiability,
observationally equivalent counter-structures, and a finite catalogue of
structural attack modes. A persuasion overlay projects each claim into
authority, expert-tone, confidence, consensus and concealment styles. Splits
enforce partitioned holdouts at the family, mechanism, attack-family,
context-shift, and paired-flip levels, and leakage-prone metadata fields are
stripped from the verifier-side view by construction
(Section~\ref{sec:benchmark}). Following our current research plan, the
literature-grounded subset and dual human audit are treated as required
completion gates rather than finished evidence.

\paragraph{Our reference verifier.} \emph{Countermodel-grounded selective verification}
(Section~\ref{sec:method}) is a four-stage pipeline that (i) parses the claim
into structured slots, (ii) builds an assumption ledger of identifying
preconditions, (iii) searches for observationally compatible counter-SCMs that
disagree with the claim on the target query, and (iv) consults Pearl-style
identification tools as supporting evidence. A fixed decision rule turns these
artifacts into one of the three verdicts plus an abstention reason and a
missing-information specification.

\paragraph{Current evidence.} Three propositions
(Section~\ref{sec:theory}) tie selective abstention, countermodel witnesses,
and oracle leakage to formal properties of the task. Empirically
(Section~\ref{sec:experiments}), the current three-seed,
ten-samples-per-family workspace snapshot supports three diagnostic findings:
(a) the reference verifier has a better OOD tradeoff than public-view heuristic
baselines ($0.891$ accuracy and $0.000$ unsafe acceptance), but this should be
read as a fixed-code sanity check rather than a saturation result; (b) a
deliberate oracle-leaking positive control reaches $1.00$ accuracy on both
dedicated attack-only splits, illustrating why public/gold partitioning must be
audited; and (c) removing \emph{abstention} preserves high IID raw accuracy
($0.856$) but collapses unidentifiable-awareness to $0.00$ on both splits. We
discuss limitations -- in particular, the still-modest benchmark scale, missing
real-grounded human audit, and prompt-protocol-specific API model baselines --
in Section~\ref{sec:experiments} and Section~\ref{sec:limitations}.

\section{Related Work}
\label{sec:related}

\paragraph{Causal benchmarks for LLMs.}
\textsc{CLadder} \cite{jin2023cladder} introduces a $10$K-sample symbolic
causal benchmark spanning Pearl's three rungs, with ground truth derived from
an oracle inference engine. \textsc{Corr2Cause} \cite{jin2024corr2cause} probes
whether LLMs can convert correlational statements into causal structure; the
authors report near-random performance and a sharp drop under perturbation.
\textsc{CRASS} \cite{frohberg2022crass} targets counterfactual conditionals,
and \textsc{CausalReasoningBenchmark}
\cite{sawarni2026causalreasoningbenchmark} separates causal identification
from numerical estimation on real datasets. We differ from these benchmarks in
three structural ways: (i) we strip oracle fields from the verifier's view by
schema construction rather than by prompt hygiene, (ii) we elevate
\textsc{unidentifiable} to a first-class supervised label tied to public
evidence, and (iii) we evaluate not just accuracy but a wise-refusal-aware
metric bundle and a deliberately leaking positive control.

\paragraph{Fact verification and real-grounding sources.}
\textsc{CHECKWHY} \cite{si2024checkwhy} studies causal fact verification with
claim-evidence-argument triplets and support/refute/not-enough-info labels,
while \textsc{SciFact} \cite{wadden2020scifact} and \textsc{HealthVer}
\cite{sarrouti2021healthver} focus on scientific and health claim
verification. Clinical evidence corpora such as Evidence Inference 2.0
\cite{deyoung2020evidenceinference} and \textsc{EvidenceOutcomes}
\cite{zhou2025evidenceoutcomes} provide treatment-comparator-outcome language,
and causal/uplift datasets such as ACIC challenges \cite{acic2019},
\textsc{RealCause} \cite{neal2021realcause}, Criteo Uplift \cite{criteo2014uplift},
\textsc{KuaiRec}/\textsc{KuaiRand} \cite{gao2022kuairec,gao2022kuairand},
and the Open Bandit Dataset \cite{saito2021openbandit} provide grounded
settings with interventions, exposure policies, or logged decisions. We use
these as target sources for the planned real-grounded subset, but their native
labels do not directly equal our
$\{\textsc{valid},\textsc{invalid},\textsc{unidentifiable}\}$ causal
identifiability contract and therefore require re-annotation.

\paragraph{Selective prediction and abstention.}
Classical selective prediction
\cite{geifman2017selective,geifman2019selectivenet} introduces a reject option
to deep classifiers. Recent work re-derives this idea for LLMs and finds that
abstention quality is sensitive to calibration
\cite{guo2017calibration,wen2025abstention}. Our task definition is closest to
this literature in spirit: the verifier explicitly outputs an abstention reason
when public evidence does not determine the query, for example observational
equivalence or missing identifying support.

\paragraph{LLM-as-judge and sycophancy.}
LLM judges show systematic biases: sycophancy
\cite{sharma2024sycophancy}, position and verbosity bias \cite{ye2024justice},
and authority effects \cite{chen2024judge}. Causal claims compound these biases
because the surface plausibility of an explanation is correlated with the
structural correctness of the underlying identification. Our persuasion
overlay is designed to expose exactly that gap.

\paragraph{Scalable oversight and debate.}
\citet{irving2018debate} and \citet{bowman2022measuring} frame oversight as a
game between informed agents and a less-informed judge. Our setting is a
specialization of that frame to causal claims, with a structured rather than
free-form transcript and an explicit abstention option for the judge.

\paragraph{Causal identification and tools.}
We build on canonical identification machinery: do-calculus and the
backdoor/frontdoor criteria \cite{pearl2009causality,shpitser2008complete},
LATE-style instrumental-variable identification \cite{imbens1994late}, and
probability-of-causation results \cite{tian2000probabilities}. Tool execution
in our pipeline uses estimators in the spirit of DoWhy
\cite{sharma2020dowhy}.

\section{Task: Selective Adversarial Causal Oversight}
\label{sec:task}

\subsection{Information Partition}
A scenario is generated as a structural causal model
$\mathcal{M} = (\mathbf{V}, \mathbf{U}, F, P(\mathbf{U}))$ over observed
variables $\mathbf{V}$ and hidden variables $\mathbf{U}$. We expose three
contracted views:

\begin{itemize}
\item \emph{Attacker view}: the gold SCM, the true DAG, hidden variable
  identities, full data, and a generator metadata bundle. The attacker uses
  this to craft a claim that is rhetorically plausible against public
  evidence but structurally constrained.
\item \emph{Verifier view}: a \texttt{PublicCausalInstance} containing only
  observed variables, public observed data, optional named role hints
  (proxy/instrument/mediator/selection), and the natural-language claim.
\item \emph{Evaluator view}: gold + verifier views, used solely to score the
  verdict and audit the witnesses.
\end{itemize}

The contract is enforced at the schema level: the
\texttt{PublicCausalInstance} dataclass strips gold-only fields such as the
true DAG, hidden variables, full SCM, full data, and gold labels, and sanitizes
public text against explicit leakage patterns before the verifier-side view is
emitted.

\subsection{Output and Label Spaces}

For a parsed claim with target $(X, Y)$ over query type
$q \in \{$association, intervention, counterfactual$\}$, the verifier outputs

\begin{itemize}
\item a verdict $\hat{y} \in \{\textsc{valid}, \textsc{invalid},
  \textsc{unidentifiable}\}$,
\item an identification status $s$ that is either \textsc{identified},
  \textsc{contradicted}, or \textsc{underdetermined}, consistent with
  $\hat{y}$,
\item a refusal reason $r$ when $\hat{y} = \textsc{unidentifiable}$,
\item an assumption ledger of typed entries
  (\textsc{supported}, \textsc{contradicted}, \textsc{unresolved}),
\item witness slots for supporting evidence, countermodels, and assumptions,
  and
\item a missing-information spec listing the unsupported assumptions and the
  evidence that would close the identification gap.
\end{itemize}

\subsection{Why the Third Label is Necessary}
\label{sec:third-label}

For three of our twelve graph families (\texttt{l1\_latent\_confounding},
\texttt{l2\_invalid\_iv}, \texttt{l3\_counterfactual\_ambiguity}), there exist
multiple SCMs $\mathcal{M}_1, \mathcal{M}_2$ that are observationally
equivalent but disagree on the target query. Pearl's causal hierarchy theorem
\cite{bareinboim2022pearl} formalizes the general case: layer-$k$ data
underdetermines layer-$(k+1)$ truth almost everywhere. We therefore treat
\textsc{unidentifiable} not as a degraded label but as the unique correct
output when public evidence does not pin down a single query value.

\section{Benchmark}
\label{sec:benchmark}

\subsection{Synthetic Core: Twelve Graph Families}

The benchmark generator instantiates each scenario from a registry of twelve
deterministic graph family blueprints (Table~\ref{tab:families}), distributed
across the three rungs of Pearl's hierarchy (four L1, five L2, three L3
families). Each family is tagged with a family-level
identifiability status (\textsc{identifiable} vs.\
\textsc{potentially\_unidentifiable}), a list of supported gold labels, and a
catalogue of attack modes that match the family's structure.

\begin{table*}[t]
\caption{Twelve graph families across Pearl's hierarchy. Each family fixes
  observable roles (treatment, outcome, context, adjuster, instrument, mediator,
  proxy, selection, latent confounder), a supported label set, and an attack
  catalogue. ``ID'' = structurally identifiable; ``PU'' = potentially
  unidentifiable.}
\label{tab:families}
\begin{center}
\begin{small}
\begin{sc}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllll}
\toprule
Rung & Family & ID/PU & Supported Labels & Primary Attack Modes \\
\midrule
L1 & latent\_confounding         & PU & invalid, unidentifiable & assoc.\ overclaim, hidden-conf.\ denial \\
L1 & selection\_bias             & PU & invalid, unidentifiable & selection-bias obfuscation \\
L1 & proxy\_disambiguation       & ID & valid, invalid          & assoc.\ overclaim \\
L1 & reverse\_causality          & PU & invalid, unidentifiable & assoc.\ overclaim \\
L2 & valid\_backdoor             & ID & valid, invalid          & invalid-adjustment, heterogeneity overgen.\ \\
L2 & valid\_iv                   & ID & valid, invalid          & heterogeneity overgen.\ \\
L2 & invalid\_iv                 & PU & invalid, unidentifiable & weak-IV-as-valid-IV, exclusion claim \\
L2 & subgroup\_heterogeneity     & PU & invalid, unidentifiable & heterogeneity overgen.\ \\
L2 & frontdoor\_partial\_meas.   & PU & valid, unidentifiable   & unidentifiable disguised as valid \\
L3 & counterfactual\_ambiguity   & PU & unidentifiable          & counterfactual overclaim \\
L3 & mediation\_abduction        & ID & valid, invalid          & function-form manipulation \\
L3 & monotonicity\_cross\_world  & PU & invalid, unidentifiable & cross-world failure, function-form \\
\bottomrule
\end{tabular}
}
\end{sc}
\end{small}
\end{center}
\end{table*}

\subsection{Persuasion Overlay}

A separate overlay layer projects each structurally-fixed claim into a
persuasion-styled variant. The taxonomy mirrors the social-science persuasion
literature \cite{zeng2024johnny} but is deliberately narrowed to five axes:
\textsc{authority}, \textsc{expert\_tone}, \textsc{confidence},
\textsc{consensus}, and \textsc{concealment}. The paper-facing primary axis
uses four of these and reports \textsc{expert\_tone} only in the
taxonomy-complete artifact, to avoid double-counting tone effects with
authority effects in the main table.

\subsection{Witnesses}

Every gold sample carries up to three structured witnesses:
a support witness (why the gold label holds under public evidence), a
countermodel witness (an observationally compatible alternative SCM that
disagrees with the claim on the target query), and an assumption witness
(which identifying assumption is unsupported or contradicted). These witnesses
are first-class evaluation objects rather than narrative decoration: they
ground the decision rule in Section~\ref{sec:method} and the qualitative
witness-faithfulness probe in Section~\ref{sec:experiments}.

\subsection{Splits and Holdouts}

The split manifest enforces non-overlapping \texttt{train}, \texttt{dev},
\texttt{test\_iid}, and \texttt{test\_ood} subsets. The holdout strategy
explicitly tracks several orthogonal axes: graph family, mechanism
(e.g.\ confounding vs.\ collider), attack-family, context-shift domain,
paired-flip pairs, plus lexical/template and variable-renaming controls inside
the generator. This goes beyond the lexical-perturbation OOD studied in prior
work \cite{jin2024corr2cause}, which we view as a useful but insufficient
stress test. Section~\ref{sec:experiments} reports the five paper-facing OOD
buckets that correspond directly to the benchmark contract: graph-family,
mechanism, attack-family, context-shift, and paired-flip OOD.

\subsection{Real-Grounded Subset}

A literature-grounded subset (target size $150$--$300$ cases) is curated from
economics, policy, epidemiology and education observational studies. Each case
records a source citation, a public-evidence summary, the
visible-vs-hidden information contract, the gold label, identifying
assumptions, and a witness note. The synthetic and real-grounded subsets share
the same claim schema, so models trained or tuned on the synthetic core can be
zero-shot evaluated on the real-grounded subset. This subset is part of the
benchmark \emph{design}; the current workspace also includes a $150$-case
human-audit package generator, but not yet dual-annotator agreement results.
Full real-grounded evaluation is therefore reserved for the human-audit phase
(see Section~\ref{sec:limitations}).

\section{Method: Countermodel-Grounded Selective Verification}
\label{sec:method}

The verifier pipeline runs four sequential stages and emits a single
\texttt{VerifierDecision} object.

\paragraph{Stage 1: Claim parsing.}
Given a public scenario and the natural-language claim, the parser extracts
treatment, outcome, query type, polarity, strength, mentioned and implied
assumptions, the rhetorical strategy (e.g.\ an adjustment-sufficiency claim
or an IV appeal), and an abstention-risk flag derived from claim-level cues.

\paragraph{Stage 2: Assumption ledger.}
The ledger initializes one entry per identifying precondition required by the
parsed claim. Status starts as \textsc{unresolved} and is updated by
later stages: tool-backed evidence may flip an entry to \textsc{supported},
contradictory evidence flips it to \textsc{contradicted}. Baseline
assumptions (\texttt{consistency}, \texttt{positivity}) are tracked but not
counted as identifying support.

\paragraph{Stage 3: Countermodel search.}
The countermodel module enumerates structurally-compatible alternative SCMs
for the public scenario and scores each on (a) the public-equivalence distance
defined in Section~\ref{sec:theory}, (b) disagreement with the claim on the
target query, and (c) which identifying assumption it triggers. The module
returns a chosen candidate, a short explanation, and a verdict suggestion.

\paragraph{Stage 4: Tool-backed adjudication.}
When no countermodel survives, a tool runner executes the appropriate
identification tools for the rung -- conditional independence and association
checks for L1, backdoor, frontdoor and IV estimators for L2, and SCM-based
counterfactual reasoning plus probability-of-necessity bounds for L3
\cite{tian2000probabilities,sharma2020dowhy}. Tool outputs are stored as a
structured trace that records which ledger entries are supported or
contradicted.

\paragraph{Decision rule.}
The four stages are combined by a fixed rule:
\begin{enumerate}
\item If a strong countermodel survives and the claim's verdict suggestion is
  \textsc{invalid}, return \textsc{invalid}.
\item Else if a countermodel survives and the candidates disagree on the
  target query, return \textsc{unidentifiable} with refusal reason
  \texttt{observational\_equivalence}.
\item Else if any identifying assumption is contradicted by tools, return
  \textsc{invalid}.
\item Else if any core identifying assumption remains unsupported or no
  primary-supporting tool record exists, return \textsc{unidentifiable} with
  a missing-identifying-support reason; when the abstention gate fires on a
  high-risk claim, we return the stricter primary-support variant.
\item Else return \textsc{valid}.
\end{enumerate}
This rule is deterministic, auditable, and is implemented in fewer than $300$
lines of Python; the closed-form schema makes selective verdicts directly
comparable across systems.

\section{Theoretical Guarantees}
\label{sec:theory}

We make three minimal claims that pin down why the task design is necessary
rather than aesthetic.

\begin{proposition}[Abstention Necessity]
\label{prop:abstention}
Let $E$ denote the public evidence and $q$ the target query. Suppose there
exist SCMs $\mathcal{M}_1, \mathcal{M}_2$ that both induce $E$ as their
observational marginal but disagree on $q$. Then any verifier that does not
admit {\normalfont\textsc{unidentifiable}} as an output produces an incorrect
verdict on
at least one of $\mathcal{M}_1, \mathcal{M}_2$.
\end{proposition}
\begin{proof}[Proof sketch]
Both SCMs are public-evidence-equivalent, so the verifier's input distribution
is identical under either generative model. Any committed verdict is therefore
the same under both, but $q(\mathcal{M}_1) \neq q(\mathcal{M}_2)$, so the
verdict is wrong under exactly one.
\end{proof}

\emph{Empirical correspondence.} Section~\ref{sec:experiments}
(\textsc{No-Abstention} ablation) instantiates this proposition: removing the
abstention head retains $0.856$/$0.462$ raw accuracy on IID/OOD but zeroes out
unidentifiable-awareness.

\begin{definition}[$\varepsilon$-public equivalence]
\label{def:public-equivalence}
Fix a public projection $\Pi(\mathcal{M})$ that contains exactly the fields
available to the verifier, and a finite public evidence map
$\phi(\Pi(\mathcal{M}))$ containing the symbolic public fields plus the
observable summary statistics used by the verifier. Two SCMs are
$\varepsilon$-public-equivalent, written
$\mathcal{M}_1 \equiv_{\Pi,\varepsilon} \mathcal{M}_2$, when their symbolic
public fields match and
$\|\phi(\Pi(\mathcal{M}_1))-\phi(\Pi(\mathcal{M}_2))\|_\infty \le
\varepsilon$ on the numeric public summaries.
\end{definition}

\begin{proposition}[Countermodel Witness Soundness]
\label{prop:witness}
Suppose a verifier exhibits an SCM $\mathcal{M}'$ such that
$\mathcal{M}' \equiv_{\Pi,\varepsilon} \mathcal{M}$ for the benchmark's
declared public evidence map, but $q(\mathcal{M}') \neq q(\mathcal{M})$ for
the target query. Then no verifier whose input is restricted to the
$\varepsilon$-public-equivalence class can uniquely identify the value of
$q$. A countermodel witness is therefore a certificate of non-identifiability
relative to $(\Pi,\phi,\varepsilon)$, not a claim about all possible evidence
outside the public contract.
\end{proposition}
\begin{proof}[Proof sketch]
The verifier's admissible input is invariant over the
$\varepsilon$-public-equivalence class, while the query value differs across
two members of that class. Any rule that commits to a unique value of $q$ from
that input alone must be wrong for at least one member. The qualitative
shuffle/delete probe in Section~\ref{sec:experiments} provides exploratory
empirical support that these witnesses are load-bearing in the implemented
pipeline.
\end{proof}

\begin{proposition}[Oracle Leakage Inflation]
\label{prop:leakage}
Let $V_{\text{public}}$ be any fixed verifier over public inputs, and let
$g \in \texttt{GOLD\_ONLY\_FIELDS}$ contain a decodable label signal. Define an
oracle-positive-control verifier $V_{\text{leak}}^\star$ that uses
$V_{\text{public}}$ when the decoded signal is unavailable and otherwise
returns the decoded label. Then
$\mathbb{E}[\text{Acc}(V_{\text{leak}}^\star)] \ge
 \mathbb{E}[\text{Acc}(V_{\text{public}})]$, with strict inequality whenever
the decoded signal corrects at least one expected error of $V_{\text{public}}$.
This is not a monotonicity statement about arbitrary verifiers with extra
fields; it is a positive-control construction.
\end{proposition}

\emph{Empirical correspondence.} Section~\ref{sec:experiments} (Leakage Study)
instantiates this proposition with a positive-control verifier; the artifact's
\texttt{supports\_warning} field operationalizes the proposition as a
benchmark-level contamination gate.

The role of \cref{prop:leakage} is operational: it justifies our leakage
study (Section~\ref{sec:experiments}) by predicting that any inflation
observed there is, by construction, leakage rather than oversight ability.

\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}
\label{sec:setup}

We report the current workspace snapshot: three seeds with ten samples per
family across all twelve graph families before split filtering. The split
builder then assigns instances to \texttt{train}, \texttt{dev},
\texttt{test\_iid}, and \texttt{test\_ood}; consequently, the API baseline table
reports the evaluated post-split counts ($n=28$ IID and $n=221$ OOD per model
after seed aggregation) rather than the pre-split generator budget. The main
benchmark runs at difficulty $0.55$ and compares the full
\emph{countermodel-grounded} verifier against public-view heuristic
judge/debate, tool-only and refusal-aware baselines. The component ablation
suite replaces one or more pipeline stages with a no-op surrogate. We also
include a dedicated attack-only leakage study under the
hidden-information-aware profile. These runs are reproducible from the current
repository state. All paper-facing tables in this draft are generated by
\texttt{python -m experiments.paper\_artifacts} from the hashed bundle recorded
in \texttt{outputs/paper\_facing/manifest.json}. They are not a saturation
study: the synthetic core is still modest, the leakage study is exploratory
because it uses ten rather than the pre-registered sixty samples per family,
and the real-grounded and human-audit phases are not yet complete. We also
run a live API LLM baseline matrix over five configured models
(\texttt{qwen2.5-7b-instruct}, \texttt{qwen2.5-14b-instruct},
\texttt{qwen2.5-72b-instruct}, \texttt{gpt-5.5} with xhigh reasoning, and
\texttt{deepseek-v4-pro} with max thinking). This matrix covers the same
three seeds and \texttt{test\_iid}/\texttt{test\_ood} splits, records prompts
and raw responses, rejects mock fallback, and produced $1245$ predictions with
zero parse errors and zero fallback records. We therefore present the empirical
results as fixed-code evidence for the benchmark contract, the reference
verifier, and the current API-backed prompt protocol, not as a final benchmark
release or a broad model-ranking claim.

\subsection{Main Benchmark}

Table~\ref{tab:main} compares the full \textsc{Countermodel-Grounded}
verifier against compact public-view heuristic baselines on \texttt{test\_iid}
and \texttt{test\_ood}. These baselines are deterministic runner
implementations named after common judge/debate prompting patterns; they are
not claimed to be strong API LLM baselines. We report the three metrics most
central to our task definition: verdict accuracy, unsafe-acceptance rate, and
wise-refusal recall. \textsc{Self-Consistency Judge} matches \textsc{CoT
Judge} exactly in this snapshot and is omitted from the compact table for
readability; the full artifact retains it.

\begin{table*}[t]
\caption{Main benchmark, current mainline snapshot (three seeds, ten
  samples per family, difficulty $0.55$). Mean $\pm$ std across seeds; full
  $95\%$ CIs, calibration metrics, and over-refusal rates are reported in
  \texttt{outputs/review\_fixed\_exp\_main\_benchmark.md}.}
\label{tab:main}
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{llccc}
\toprule
System & Split & Verdict Acc.\ & Unsafe Accept & Wise Refusal Recall \\
\midrule
\input ../../outputs/paper_facing/main_table_rows.tex
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\end{table*}

Three observations matter. First, \textsc{Direct Judge} exhibits exactly the
failure mode that motivates the task: extremely high unsafe acceptance
($0.89$ IID, $0.97$ OOD) under public evidence constraints. Second,
\textsc{CoT Judge}, \textsc{Debate Baseline}, and the explicit
\textsc{Refusal-Aware} baseline drive unsafe acceptance to zero, but do so by
over-refusing: on \texttt{test\_ood}, their over-refusal rates are
$0.850$, $0.677$, and $0.557$, respectively, versus $0.010$ for the full
system. Third, the near-perfect IID score should be read as a reference-verifier
sanity check on the synthetic contract, not as evidence that generic LLMs have
solved causal oversight. The stronger claim is the tradeoff shape: the full
pipeline maintains zero unsafe acceptance while preserving OOD wise-refusal
recall and substantially improving raw verdict accuracy over public-view
heuristic baselines.

\subsection{Strong API LLM Baselines}
\label{sec:llm-baselines}

Table~\ref{tab:llm-baselines} reports the completed live API baseline matrix.
Each model receives the same public-view prompt used by the API runner: a
compact public scenario, the natural-language claim, and an instruction to
return only a JSON verdict in
$\{\textsc{valid}, \textsc{invalid}, \textsc{unidentifiable}\}$. All gold-only
fields, countermodel witnesses, hidden variables, and attacker rationales are
excluded from the prompt by the same leakage guard used in the smoke tests. The
matrix is small and prompt-protocol-specific, but fully auditable: 30
model/seed/split jobs, $1245$ predictions, raw-response logs, model settings,
prompt hashes, and zero fallback or parse-error records. The $n$ column reports
post-split evaluated instances aggregated over the three seeds.

\begin{table*}[t]
\caption{Live API LLM baseline matrix. Qwen rows use DashScope
  chat-completions; GPT-5.5 uses the Responses API with xhigh reasoning;
  DeepSeek uses chat-completions with max thinking. Results are generated from
  \texttt{outputs/llm\_baseline\_matrix/llm\_baseline\_aggregated\_metrics.json};
  $n$ is the post-split evaluated count aggregated over three seeds.}
\label{tab:llm-baselines}
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{llccccc}
\toprule
Model & Split & $n$ & Acc.\ & Macro-F1 & Unsafe Accept & Wise Refusal Recall \\
\midrule
\input ../../outputs/paper_facing/llm_baseline_table_rows.tex
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\end{table*}

The strongest raw OOD accuracy among these API-backed baselines is GPT-5.5
xhigh at $0.545$, with unsafe acceptance $0.071$ and wise-refusal recall
$0.608$. DeepSeek v4 with max thinking reaches similar IID accuracy
($0.715$) but weaker OOD selective behaviour ($0.461$ accuracy and $0.218$
wise-refusal recall). The Qwen family shows the expected size-sensitive but
non-monotone tradeoff: Qwen2.5-14B has the best OOD accuracy among the Qwen
models ($0.523$) and zero unsafe acceptance, while Qwen2.5-72B has higher
wise-refusal recall ($0.876$) but still over-refuses substantially. The gap
between these public-view LLM baselines and the countermodel-grounded
reference verifier is not evidence that generic LLMs cannot reason causally, nor
a definitive leaderboard result; it is evidence that this public/gold partition
and three-way selective prompt protocol surfaces failure modes that are hidden
by closed-world causal QA scoring.

\subsection{Component Ablation: What Drives Wise Refusal?}

Table~\ref{tab:ablation} extends the ablation to four pipeline components,
including a \textsc{No-Abstention} variant that disables the third-label
output entirely. Two patterns are especially important. First,
\textsc{No-Abstention} preserves high IID raw accuracy ($0.856$) but zeroes
out unidentifiable-awareness on both splits, which is exactly the failure mode
the selective task is designed to surface. Second, removing the countermodel
stage hurts both OOD verdict accuracy and OOD unidentifiable-awareness
($0.504$ and $0.591$). The current snapshot is not a strict monotone ranking
across every metric: \textsc{No-Tools} exceeds the full system on OOD raw
accuracy ($0.911$ vs.\ $0.891$) and unidentifiable-awareness
($0.964$ vs.\ $0.825$), but does so with substantially higher over-refusal
($0.150$ vs.\ $0.010$) and lower wise-refusal precision
($0.877$ vs.\ $0.989$). This result weakens any claim that tool-backed
adjudication is uniformly beneficial in the current synthetic snapshot. We
therefore interpret the full system as a conservative reference verifier with a
better precision/over-refusal balance, not as uniformly best on every scalar
metric.

\begin{table}[t]
\caption{Component ablation. Unidentifiable-awareness is the task-critical
  third-label measure; the full artifact includes precision, over-refusal,
  calibration, and countermodel coverage.}
\label{tab:ablation}
\vskip 0.05in
\begin{center}
\begin{small}
\begin{sc}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{llcc}
\toprule
System & Split & Acc.\ & Unident.\ Awareness \\
\midrule
\input ../../outputs/paper_facing/ablation_table_rows.tex
\bottomrule
\end{tabular}
}
\end{sc}
\end{small}
\end{center}
\end{table}

\subsection{Leakage Study}

The current leakage study is explicitly exploratory because it uses only ten
samples per family. We therefore present it as a positive-control sanity check
that the public/gold contract behaves as Proposition~\ref{prop:leakage}
predicts, rather than as a definitive estimate of contamination magnitude.
Full leakage evidence requires increasing samples\_per\_family to at least
$60$.

We instantiate two verifiers on a dedicated attack-only benchmark with the
hidden-information-aware attack profile: a clean public verifier and a
deliberately oracle-leaking control that decodes benchmark-answer supervision
from a protocol-violating public metadata partition. Per \cref{prop:leakage},
any positive gap is
\emph{by construction} leakage inflation. Table~\ref{tab:leakage} shows a gap
of $+0.421$ on \texttt{test\_iid} and $+0.155$ on \texttt{test\_ood}. Both
paired tests support the warning in this small run (bootstrap
$p=0.0395/0.0015$, McNemar $p=0.0156/0.0000$ before the table's rounding), but
the protocol notice still prevents treating this as formal leakage evidence
until the sample count is increased. ECE drops from $0.400$ to $0.000$ on
\texttt{test\_iid} and from $0.077$ to $0.000$ on \texttt{test\_ood},
illustrating how leaked supervision can masquerade as calibration.

\begin{table}[t]
\caption{Leakage study. Clean public verifier vs.\ an oracle-positive control
  with protocol-violating public metadata. Any positive gap is leakage inflation
  (\cref{prop:leakage}). Numbers from the accompanying
  \texttt{outputs/review\_fixed\_exp\_leakage\_study.md} artifact.}
\label{tab:leakage}
\vskip 0.05in
\begin{center}
\begin{small}
\begin{sc}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lccccc}
\toprule
Split & Clean Acc. & Leaking Acc. & $\Delta$ & Clean 95\% CI & Leaking 95\% CI \\
\midrule
\input ../../outputs/paper_facing/leakage_table_rows.tex
\bottomrule
\end{tabular}
}
\end{sc}
\end{small}
\end{center}
\end{table}

\subsection{Adversarial Robustness}

Table~\ref{tab:adv} reports robustness across four attack-strength regimes:
\textsc{weak}, \textsc{medium}, \textsc{strong}, and a leak-aware
hidden-information profile. On \texttt{test\_ood}, the verifier stays
in a narrow $0.845$ band across all four regimes while maintaining zero invalid
acceptance. The \texttt{test\_iid} picture is less stable: the
hidden-information-aware regime drops to $0.579 \pm 0.417$, making it the
hardest current IID stress case. Because the weak/medium/strong rows are nearly
identical in this snapshot, we treat this section as exploratory stress
testing rather than a headline robustness claim.

\begin{table}[t]
\caption{Adversarial robustness. Mean $\pm$ std across seeds; full results are
  reported in the accompanying \texttt{exp\_adversarial\_robustness.md}
  artifact.}
\label{tab:adv}
\vskip 0.05in
\begin{center}
\begin{small}
\begin{sc}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{llcc}
\toprule
Strength & Split & Acc.\ & Wise Refusal Recall \\
\midrule
\input ../../outputs/paper_facing/adversarial_table_rows.tex
\bottomrule
\end{tabular}
}
\end{sc}
\end{small}
\end{center}
\end{table}

\subsection{OOD Buckets}

The current OOD artifact evaluates the five buckets defined by the benchmark
contract: graph-family, mechanism, attack-family, context-shift, and
paired-flip OOD. The graph-family and mechanism buckets are saturated in this
small snapshot ($1.000$ accuracy), so they are best read as contract smoke tests
rather than convincing robustness evidence. Attack-family ($0.820$) and
context-shift ($0.883$) are more informative, while paired-flip drops sharply
($0.439$) and has the largest macro-F1 gap ($0.423$), making it the clearest
current stress case for structural disambiguation. The current significance
artifact reports uninformative $p=1$ rows for these bucket gaps, so we avoid
strong significance claims until the OOD estimand and sample count are upgraded.

\begin{table}[t]
\caption{OOD bucket breakdown for the full verifier. The rows are generated
  from \texttt{outputs/paper\_facing/ood\_table\_rows.tex}; CIs are bootstrap
  $95\%$ intervals over seed means.}
\label{tab:ood}
\vskip 0.05in
\begin{center}
\begin{small}
\begin{sc}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lccccc}
\toprule
Bucket & Acc.\ & Macro-F1 & Unsafe Accept & Wise Refusal Recall & Acc.\ 95\% CI \\
\midrule
\input ../../outputs/paper_facing/ood_table_rows.tex
\bottomrule
\end{tabular}
}
\end{sc}
\end{small}
\end{center}
\end{table}

\subsection{Witness Faithfulness (qualitative)}

We run a small witness-faithfulness probe by shuffling or deleting the
countermodel witness in the verifier output and re-scoring. When the
countermodel witness is shuffled across instances, the system's
unidentifiable-awareness collapses on the families that rely on
observational-equivalence reasoning
(\texttt{l1\_latent\_confounding\_family},
\texttt{l3\_counterfactual\_ambiguity\_family}). This indicates that the
witness is load-bearing rather than ornamental. A full quantitative study
(witness shuffle/delete with paired-bootstrap
$95\%$ CIs across all twelve families) is reserved for future work.

\section{Discussion and Limitations}
\label{sec:limitations}

\paragraph{What we are not claiming.}
We are \emph{not} claiming a state-of-the-art result on a saturated benchmark,
nor that LLM judges are uniformly bad at causal reasoning. The claim is
structural and diagnostic: once the task enforces a public/gold contract and
rewards correct abstention, the measured failure modes differ from closed-world
causal QA. The current reference verifier is an auditable yardstick for that
contract, not a validated general-purpose causal auditor.

\paragraph{Limitations.}
The synthetic core is, by construction, a model of the world rather than the
world itself. The real-grounded subset is small, and full human dual-annotation
is in progress. The current mainline snapshot uses ten samples per family, so
bucket-level estimates can still have wide CIs; the dedicated leakage study is
explicitly exploratory; saturated graph/mechanism OOD buckets should not be
treated as robust generalization evidence; and the hidden-information-aware IID
robustness slice is not yet stable enough to support strong quantitative
claims. The API-backed model evidence now includes a completed five-model
baseline matrix with raw responses and fallback rejection, but it is still only
one prompt protocol over the current synthetic snapshot; larger budgets, more
model families, completed human audit, and pre-registered model/version freezes
will be needed before any broad statement about absolute performance is
defensible. The existing transfer matrix is likewise a
pilot: it shows that family-swapped attacker/verifier profiles are testable
inside the benchmark contract, but it is still too small to support
model-family conclusions. We document these caveats in the artifact and in
this paper.

\paragraph{Completion gates.}
Before release, three gates are required: (i) closing the real-grounded subset
with dual annotation and Cohen's $\kappa$ on the four primary audit fields;
(ii) scaling the current synthetic snapshot and transfer pilot into a
full-budget cross-model study across verifier families and sizes; and (iii)
extending the completed API baseline matrix to a pre-registered model/version
freeze with more closed and open model families.

\section{Conclusion}

Selective adversarial causal oversight is a useful unit of analysis for
LLM-based causal evaluation. By modelling information asymmetry, treating
\textsc{unidentifiable} as a first-class label, partitioning public from gold
at the schema level, and grounding the reference verifier in countermodel
witnesses, the current prototype exposes a measurable tradeoff between unsafe
acceptance and wise refusal. The next step is validity, not presentation: more
samples, real-grounded human-audited cases, stronger surface-bias checks, and
cross-model transfer.

\section*{Software and Data}
The repository contains the benchmark generator, verifier pipeline, experiment
runners, fixed-code result artifacts, and regression tests used for the current
snapshot. The current paper-facing reproducibility package is
\texttt{outputs/paper\_facing/}: its manifest records source artifact hashes,
configuration, seeds, generated table rows, poster metrics, and the available
API smoke artifact. The completed LLM baseline matrix is stored in
\texttt{outputs/llm\_baseline\_matrix/} with a manifest, aggregate JSON,
summary CSV, raw JSONL predictions, prompt hashes, model settings, and raw
responses. Each synthetic experiment emits a configuration JSON, a seed list,
raw per-sample predictions, an aggregated metrics file, $95\%$ CIs,
paired-significance tests, and a markdown summary. The leakage study
additionally emits a leakage-warning field that downstream consumers can use
as a contamination check. The human-audit gate remains explicitly pending
rather than backfilled.

\section*{Impact Statement}
This paper presents work whose goal is to advance the field of Machine
Learning. There are many potential societal consequences of our work, none
which we feel must be specifically highlighted here. We note in particular
that our adversarial framing is intended for evaluation and oversight rather
than deployment: the persuasion overlay is a stress-test artifact for
verifiers, not a recipe for deceptive content generation. The leakage decoder
is isolated as a positive-control diagnostic so that contamination risks are
measurable rather than hidden.

\nocite{pearl2009causality,pearl2018book,bareinboim2022pearl}
\nocite{jin2023cladder,jin2024corr2cause,frohberg2022crass}
\nocite{geifman2017selective,geifman2019selectivenet}
\nocite{sharma2024sycophancy,chen2024judge,ye2024justice}
\nocite{irving2018debate,bowman2022measuring}
\nocite{guo2017calibration,wen2025abstention}
\nocite{tian2000probabilities,sharma2020dowhy,imbens1994late,shpitser2008complete}
\nocite{zeng2024johnny,lanham2023measuring}
\nocite{si2024checkwhy,sawarni2026causalreasoningbenchmark,deyoung2020evidenceinference}
\nocite{zhou2025evidenceoutcomes,acic2019,neal2021realcause,criteo2014uplift}
\nocite{gao2022kuairec,gao2022kuairand,saito2021openbandit,wadden2020scifact,sarrouti2021healthver}

\bibliography{example_paper}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Schema Excerpts}
\label{app:schema}

The following is an abridged view of the leakage-control contract. The full
\texttt{benchmark/schema.py} file contains additional sanitizers and
serialization helpers; only the contract-level constants are reproduced here.

\begin{verbatim}
GOLD_ONLY_FIELDS = (
    "true_dag", "hidden_variables", "true_scm",
    "full_data", "ground_truth", "gold_label",
)

VERIFIER_VISIBLE_FIELDS = (
    "scenario_id", "description", "variables", "proxy_variables",
    "selection_mechanism", "observed_data", "data",
    "causal_level", "sanitized_public_metadata",
)

class VerdictLabel(str, Enum):
    VALID = "valid"
    INVALID = "invalid"
    UNIDENTIFIABLE = "unidentifiable"

class IdentificationStatus(str, Enum):
    IDENTIFIED = "identified"
    CONTRADICTED = "contradicted"
    UNDERDETERMINED = "underdetermined"
\end{verbatim}

\section{Decision Rule (Pseudocode)}
\label{app:decision}

\begin{verbatim}
def decide_verdict(parsed, ledger, cm, tool_trace):
    # Stage 1: strong invalid countermodel
    if cm.found and cm.suggests_invalid:
        return INVALID

    # Stage 2: observational equivalence + query disagreement
    if cm.found and cm.query_disagreement:
        return UNIDENTIFIABLE(reason="observational_equivalence")

    # Stage 3a: contradicted assumption
    if any(e.contradicted for e in ledger):
        return INVALID

    # Stage 3b: missing or unsupported identification
    if any(e.unsupported for e in ledger) \
       or not tools_support_claim(tool_trace) \
       or (needs_explicit_support and no_supported_id):
        return UNIDENTIFIABLE(reason="missing_identifying_support")

    # Stage 3c: abstention gate on high-risk valid claims
    if abstention_gate_reasons(parsed, ledger, tool_trace):
        return UNIDENTIFIABLE(reason="missing_primary_identifying_support")

    # Stage 4: support
    return VALID
\end{verbatim}

\section{Persuasion Overlay Examples}
\label{app:persuasion}

A claim with structural label \textsc{unidentifiable} can be projected into
five persuasion styles. The taxonomy-complete artifact includes
\textsc{expert\_tone}; the paper-facing primary axis collapses to the four
remaining styles to avoid double-counting tone effects.

\begin{small}
\begin{itemize}
\item \textsc{Authority}: ``As established by the foundational paper of the
  field, the effect of $X$ on $Y$ is positive.''
\item \textsc{Confidence}: ``It is unambiguous, on any reading of the data,
  that $X$ causes $Y$.''
\item \textsc{Consensus}: ``Every recent meta-analysis agrees that $X$
  raises $Y$.''
\item \textsc{Concealment}: omits any mention of the latent confounder or
  selection mechanism that the public evidence cannot resolve.
\item \textsc{Expert tone}: heavy technical jargon without a corresponding
  identification argument.
\end{itemize}
\end{small}

\section{Reproducibility Notes}
\label{app:repro}

Each experiment runner emits a fixed bundle of artifacts:

\begin{verbatim}
<run>_config.json
<run>_seed_list.json
<run>_raw_predictions.jsonl
<run>_aggregated_metrics.json
<run>_ci.json
<run>.md   (paper-facing markdown summary)
\end{verbatim}

The aggregator reports seed-mean estimates and paired bootstrap $95\%$ CIs;
the significance file applies Holm-Bonferroni correction within and across
splits. The leakage runner additionally writes a
\texttt{supports\_warning} field that downstream consumers can use as a
hard gate on contamination, in line with \cref{prop:leakage}. The command
\texttt{python -m experiments.paper\_artifacts} freezes the paper-facing
manifest and LaTeX rows in \texttt{outputs/paper\_facing/}. The command
\texttt{python -m experiments.exp\_api\_baseline\_smoke.run} performs the
API-smoke path and rejects mock fallback by default; the command
\texttt{python -m experiments.exp\_llm\_baseline\_matrix.run --reuse-existing}
rebuilds the full LLM baseline manifest and aggregate tables from completed
per-job JSON outputs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
